Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2022-12-20 src/tex/ms.tex
[showyourwork] overleaf sync

%% References
\usepackage[%
    style=numeric-comp,%
    sorting=none,%
    maxnames=20,%
    url=false,%
]{biblatex}
\addbibresource{bib.bib}

@@

Consequently, the conditional probability to observe a diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_\ell$, given a hidden involvement state $\mathbf{X}=\boldsymbol{\xi}_k$ is a matrix $\mathbf{B}$ made up of products of terms from the table above:
%
\begin{equation}
    \mathbf{B} = \left( B_{k\ell} \right) = \prod_{v=1}^V P\left( Z_v = \zeta_{\ell v} \mid X_v[t_\text{D}] = \xi_{kv} \right)
\end{equation}
%
If multiple diagnostic modalities $\mathcal{O} \in \{ \text{MRI}, \text{CT}, \ldots \}$ are used, one can compute an observation matrix $\mathbf{B}^\mathcal{O}$ for each of these modalities and combine them via an outer product into a joint observations matrix
%
\begin{equation}
    \mathbf{B} = \mathbf{B}^\text{MRI} \otimes \mathbf{B}^\text{CT} \otimes \ldots
\end{equation}
%
We define the time $t=0$ to be the moment just before a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patients with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a late T-category tumor. We can use this information by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P_\text{early}\left( t_\text{D} \right)$ for early T-category patients (T1 \& T2) and $P_\text{late}\left( t_\text{D} \right)$ for late T-category patients (T3 \& T4).

The main task of of personalizing the \gls{ctv-n} definition is now to infer the probability of each of the possible states $\boldsymbol{\xi}_k$ given a diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_\ell$. Using Bayes' theorem, we get
%
\begin{equation}
    P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) = \frac{P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_k \right) P\left( \boldsymbol{\xi}_k \right)}{\sum_{r=1}^{2^V} P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_r \right) P\left( \boldsymbol{\xi}_r \right) }
\end{equation}
%
The model we are describing here will yield an estimate of $P\left( \boldsymbol{\xi}_k \right)$ and in the next subsection we will briefly explain how this probability distribution can be efficiently parameterized.

@@

The term $P \big( X_v = \xi_{iv} \mid \left\{ X_r = \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \big)$ in \cref{eq:transition_prob} implicitly depends on how we parameterize the arcs of the underlying \gls{dag}. If we use the graph displayed in \cref{fig:graph_with_obs}), we can write down a conditional probability table for the probability of $X_3$ being healthy or metastatic, based on the state of the parent \gls{lnl} $\operatorname{pa}(3) = 2$:

@@

\printbibliography[%
    title={References},%
]

>>>>

@@

Consequently, the conditional probability to observe a diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_\ell$, given a hidden involvement state $\mathbf{X}=\boldsymbol{\xi}_k$ is a matrix $\mathbf{B}$ made up of products of terms from the table above:
%
\begin{equation} \label{eq:transition_matrix}
    \mathbf{B} = \left( B_{k\ell} \right) = \prod_{v=1}^V P\left( Z_v = \zeta_{\ell v} \mid X_v[t_\text{D}] = \xi_{kv} \right)
\end{equation}
%
If multiple diagnostic modalities $\mathcal{O} \in \{ \text{MRI}, \text{CT}, \ldots \}$ are used, one can compute an observation matrix $\mathbf{B}^\mathcal{O}$ for each of these modalities and combine them via an outer product into a joint observations matrix
%
\begin{equation}
    \mathbf{B} = \mathbf{B}^\text{MRI} \otimes \mathbf{B}^\text{CT} \otimes \ldots
\end{equation}
%
We define the time $t=0$ to be the moment just before a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patients with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a late T-category tumor. We can use this information by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_D \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_D \mid \text{late} \right)$ for late T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions, mainly because they have a plausible shape for this purpose and only a single parameter.
%
\begin{equation}
    P \left( t = t_D \mid \text{T-cat.} \right) = \operatorname{B}(t_\text{max},p_\text{T-cat.})
\end{equation}
%
We will use as a latest time-step $t_\text{max} = 10$.

\subsection{The Likelihood Function}
\label{subsec:previous_work:likelihood}

Using the definitions up to this point, we can compute a vector of likelihoods for every possible diagnosis:
%
\begin{equation}
    \boldsymbol{\mathcal{L}} = \sum_{t=0}^{t_{max}} \left[ \boldsymbol{\pi} \cdot \mathbf{A}^t \cdot \mathbf{B} \right] \cdot P \left( t \mid \text{T-cat.} \right)
\end{equation}
%
This likelihood implicitly depends on how we parameterize the arcs of the \gls{dag} underlying the model and the parameterization of the distribution over diagnosis times. If we use the graph displayed in \cref{fig:graph_with_obs}), we can write down a conditional probability table for the probability of \gls{lnl} $X_3$ being healthy or metastatic, based on the state of the parent \gls{lnl} $\operatorname{pa}(3) = 2$:

Similar tables can be written for all other \glspl{lnl}. The terms in these conditional probability tables can then be used to compute the transition probability between to particular states $\boldsymbol{\xi}_i$ and $\boldsymbol{\xi}_j$ as in \cref{eq:transition_prob}. Computing this quantity for all possible transitions subsequently yields the transition matrix $\mathbf{A}$ as described in \cref{eq:transition_matrix}.

Together with the parametrizations of the distributions over the diagnosis time, the parameters that make up the transition matrix $\mathbf{A}$ comprises the set of model parameters
%
\begin{equation}
    \boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{late} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
\end{equation}
%
The main task of of personalizing the \gls{ctv-n} definition is now to infer the probability of each of the possible states $\boldsymbol{\xi}_k$ given a diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_\ell$. Using Bayes' theorem, we get
%
\begin{equation}
    P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) = \frac{P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_k \right) P\left( \boldsymbol{\xi}_k \right)}{\sum_{r=1}^{2^V} P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_r \right) P\left( \boldsymbol{\xi}_r \right) }
\end{equation}
%
The model we are describing here will yield an estimate of $P\left( \boldsymbol{\xi}_k \right)$ that depends on the parameters used to define the transition matrix $\mathbf{A}$ and the prior over the diagnosis time $P \left( t \mid \text{T-cat.} \right)$.

@@

@@

\bibliography{bib}
